---
title: "STAT-420 Data Analysis Project"
output: 
  html_document:
    toc: yes
    toc_depth: 2
    theme: simplex
---

###*Oh! The agony of an airline statistician*
####8/4/2017

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
options(scipen = 1, digits = 4, width = 80)
```

## Introduction

[This linked repository](https://github.com/ryusko2/stat420-analysis) contains our working files for this data analysis project.

Contributing team members include:

* dakline2@illinois.edu ~ Doug Kline
* anushav3@illinois.edu ~ Anusha Varadharajan
* ryusko2@illinois.edu  ~ Ryan Yusko

We have selected a dataset representing all US domestic flights from 1990 to 2009:

[http://academictorrents.com/details/a2ccf94bbb4af222bf8e69dad60a68a29f310d9a](http://academictorrents.com/details/a2ccf94bbb4af222bf8e69dad60a68a29f310d9a)

A summary of the raw data is presented here:
  
| Short Name             | Type    | Description                                                                                              |
|------------------------|---------|----------------------------------------------------------------------------------------------------------|
| Origin                 | String  | Three letter airport code of the origin airport                                                          |
| Destination            | String  | Three letter airport code of the destination airport                                                     |
| Origin City            | String  | Origin city name                                                                                         |
| Destination City       | String  | Destination city name                                                                                    |
| Passengers             | Integer | Number of passengers transported from origin to destination                                              |
| Seats                  | Integer | Number of seats available on flights from origin to destination                                          |
| Flights                | Integer | Number of flights between origin and destination (multiple records for one month, many with flights > 1) |
| Distance               | Integer | Distance (to nearest mile) flown between origin and destination                                          |
| Fly Date               | Integer | The date (yyyymm) of flight                                                                              |
| Origin Population      | Integer | Origin city's population as reported by US Census                                                        |
| Destination Population | Integer | Destination city's population as reported by US Census                                                   |

We hope to explore interactions and possibly model predictions for the response variable(s) `Passengers`, `Seats`, and/or `Flights`.  We have already observed high correlation between `Flights` and both `Seats` and `Passengers`, so we may not need all three of these variables in our model.  This correlation is expected because, as more flights are flown between destinations, there are more seats available, and only a certain number of passengers may fly in those seats.

We have also considered confirming busy travel months by analyzing the response to `date`.

The raw data was cleaned, and several potential predictor variables have been added:

| Predictor     | Coded as                    | Explanation                                             |
|---------------|-----------------------------|---------------------------------------------------------|
| `origin_regn` | *See region codes, below    | Region where flight originated                          |
| `origin_subr` | *See subregion codes, below | Subregion where flight originated                       |
| `dest_regn`   | *See region codes, below    | Destination region                                      |
| `dest_subr`   | *See subregion codes, below | Destination subregion                                   |
| `intra_regn`  | TRUE | FALSE                | Did flight takeoff and land in same region?             |
| `intra_subr`  | TRUE | FALSE                | Did flight takeoff and land in same subregion?          |
| `month`       | 1:12                        | Integer representation of month when flight(s) occurred |
| `month.chr`   | {"Jan", "Feb", ... , "Dec"} | String representation of month when flight(s) occurred  |

Regions as defined by the [U.S. Census Bureau](https://www2.census.gov/geo/pdfs/maps-data/maps/reference/us_regdiv.pdf) are coded as follows:

| Region (code)  | Subregions Included                                    |
|----------------|--------------------------------------------------------|
| Northeast (NE) | New England, Middle Atlantic                           |
| Midwest (MW)   | East North Central, West North Central                 |
| South (S)      | South Atlantic, East South Central, West South Central |
| West (W)       | Mountain, Pacific                                      |

Subregions as defined by the [U.S. Census Bureau](https://www2.census.gov/geo/pdfs/maps-data/maps/reference/us_regdiv.pdf) are coded as follows:

| Subregion (code)         | States Included*                    |
|--------------------------|-------------------------------------|
| New England (NE)         | CT, ME, MA, NH, RI, VT              |
| Middle Atlantic (MA)     | NJ, NY, PA                          |
| East North Central (ENC) | IL, IN, MI, OH, WI                  |
| West North Central (WNC) | IA, KS, MN, MO, NE, ND, SD          |
| South Atlantic (SA)      | DE, DC*, FL, GA, MD, NC, SC, VA, WV |
| East South Central (ESC) | AL, KY, MS, TN                      |
| West South Central (WSC) | AR, LA, OK, TX                      |
| Mountain (M)             | AZ, CO, ID, MT, NV, NM, UT, WY      |
| Pacific (P)              | AK, CA, HI, OR, WA                  |

\*Although Washington DC is not technically a state, it is home to two airports included in this dataset: DCA (Ronald Reagan Washington National), and IAD (Washington Dulles International).  Washington DC is in the *South Atlantic* subregion of the *South* region.

A random sample of our [cleaned] data is provided below:

```{r warning = FALSE, cache = TRUE}
#load data
flight_edges = readr::read_csv("flight_edges.csv", col_types = "icccccccccclliiiiiiicii")

#remove first column
flight_edges = flight_edges[,2:ncol(flight_edges)]

#Some months have flights flown, but no seats.  These could represent private or military
#flights.  In either case, from an airline perspective, we are not interested in
#these flights.  They represent ~ 10% of the data.  
flight_data = subset(flight_edges, flight_edges$seats > 0)
rm(flight_edges)

#sample some data, removing columns containing `city`, ICAOs, and the raw `date`
knitr::kable(flight_data[runif(5, min = 1, max = nrow(flight_data)),-c(1:2, 3, 7, 17)])
```

```{r message = FALSE, echo = FALSE}
#these libraries are required for bptest() and shapiro.test, in diagnostics function
library(lmtest)
library(MASS)

#diagnostics function
diagnostics = function(model, pcol = "grey", lcol = "dodgerblue", alpha = 0.05, plotit = TRUE, testit = TRUE) {
  if(plotit == TRUE) {
    par(mfrow = c(1,2))
    plot(fitted(model), resid(model),
         xlab = "Fitted",
         ylab = "Residuals",
         main = "Fitted vs. Residuals",
         col = pcol)
    abline(h = 0, col = lcol, lwd = 2)
    qqnorm(resid(model), col = pcol)
    qqline(resid(model), col = lcol)
  }
  if(testit == TRUE) {
    shapiro_wilk = shapiro.test(resid(model))
    breusch_pagan = bptest(model)
    list(shapiro_wilk = shapiro_wilk, breusch_pagan = breusch_pagan)
  }
}
```

## Methods

Based on a recommendation overheard in help sessions, we decided to start modeling with a subset of 10,000 observations, without any restrictions or geographical limitation.

```{r}
#full 3.2M observations currently stored in `flight_data`
sub_obs = 10000
flight_sub = flight_data[sample(1:nrow(flight_data), sub_obs), ]
```

Right off the bat, we were curious to see whether `passengers` could be modeled as a response of `month` as a factor variable, verifying high-traffic seasons of flying.

```{r}
month_model = lm(passengers ~ month.chr, data = flight_sub)
summary(month_model, data = flight_sub)
```

Even though the p-value of this regression (`r pf(summary(month_model)$fstatistic[1], summary(month_model)$fstatistic[2], summary(month_model)$fstatistic[3], lower.tail = FALSE)`) is significant, the simple representation did not prove very fruitful at all, with an adjusted $R^2$ of `r summary(month_model)$adj.r.squared`.  Interestingly, some of the months for which we would expect to see higher travel did have *relatively* lower p values.

To get a better idea of how potential predictor variables could be reacting, we took a look at a `pairs` plot:

```{r fig.height=8, fig.width=10, cache = TRUE}
#remove non-numerical and non-interesting columns
pairs(flight_sub[, -c(1:12, 17, 20)], col = "dodgerblue")
```

